---
title: "Injury_Data"
output: html_document
author: "Jordyn Maurer"
date: "2025-09-29"
---

This file involves the combination of 8 injury datasets. Unnecessary columns and injuries such as illnesses, allergies, bug bites, etc. were removed from the final combined dataset.The final dataset can be found in the github repo or Google Drive as "injury_data.csv".  

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/jordy/OneDrive/Documents/Coach_Scott/Injuries")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r setup, include=FALSE}
# loading necessary libraries
library(dplyr)
library(readr)
library(lubridate)
```


```{r setup, include=FALSE}
# replace with user's personal directory
setwd("C:/Users/jordy/OneDrive/Documents/Coach_Scott/Injuries")

# reading in injuries
aug_inj <- read.csv("Injury Report August.csv")
sep_inj <- read.csv("Injury Report September.csv")
oct_inj <- read.csv("October Injury Report.csv")
nov_inj <- read.csv("November Inj.csv")
jan_inj <- read.csv("Jan_Injuries.csv")
march_inj <- read.csv("March_Injuries.csv")
april_inj <- read.csv("April Updated.csv")
may_inj <- read.csv("May_Injuries.csv")
```


# 1. Cleaning the Individual Datasets Prior to Combination
```{r}
# August: Removing unnecessary columns
aug_inj_clean <- aug_inj %>%
  select(-Teams, -ProblemReportedBy, -IncidentReportCreatedBy, -DateOfReturn, -CurrentStatus, -ReportedDate, -IncidentReportCreatedOnDate)

# August: Turning the 'ProblemDate' variable into the proper date format
aug_inj_clean$ProblemDate <- mdy(substr(aug_inj_clean$ProblemDate, 5, 15))
str(aug_inj_clean$ProblemDate)

# September: Removing unnecessary columns
sep_inj_clean <- sep_inj %>%
  select(-Teams, -ProblemReportedBy, -IncidentReportCreatedBy, -DateOfReturn, -CurrentStatus, -ReportedDate, -IncidentReportCreatedOnDate) %>%
  rename(ID = X)

# September: Turning the 'ProblemDate' variable into the proper date format
sep_inj_clean$ProblemDate <- mdy(substr(sep_inj_clean$ProblemDate, 5, 15))
str(sep_inj_clean$ProblemDate)

# October: Removing unnecessary columns
oct_inj_clean <- oct_inj %>%
  select(-Teams, -ProblemReportedBy, -IncidentReportCreatedBy, -DateOfReturn, -CurrentStatus, -ReportedDate, -IncidentReportCreatedOnDate)

# October: Turning the 'ProblemDate' variable into the proper date format
oct_inj_clean$ProblemDate <- mdy(substr(oct_inj_clean$ProblemDate, 5, 15))
str(oct_inj_clean$ProblemDate)

# November: Removing unnecessary columns and Renaming column for consistency across datasets
nov_inj_clean <- nov_inj %>%
  select(-ProblemReportedBy, -IncidentReportCreatedBy, -DateOfReturn, -CurrentStatus, -ReportedDate, -IncidentReportCreatedOnDate) %>%
  rename(ID = Teams)

# November: Turning the 'ProblemDate' variable into the proper date format
nov_inj_clean$ProblemDate <- mdy(substr(nov_inj_clean$ProblemDate, 5, 15))
str(nov_inj_clean$ProblemDate)

# January: Removing unnecessary columns
jan_inj_clean <- jan_inj %>%
  select(-Expected.Return.Date, -Actual.Return.Date, -Days.Out, -Problem.Status, -Reported.Date)

# January: Renaming columns for consistency across datasets
jan_inj_clean <- jan_inj_clean %>%
  rename(ID = ID.) %>%
  rename(ProblemDate = Problem.Date) %>%
  rename(InjuryType = Condition) %>%
  rename(BodyPart = Body.Part) %>%
  rename(AffectedArea = Affected.Area) %>% rename(Side = Problem.Side) %>%
  rename(BodyPartName = Problem.Sites)

# January: Turning the 'ProblemDate' variable into the proper date format
jan_inj_clean$ProblemDate <- mdy_hm(jan_inj_clean$ProblemDate)
jan_inj_clean$ProblemDate <- as.Date(jan_inj_clean$ProblemDate)
str(jan_inj_clean$ProblemDate)

# March: Removing unnecessary columns and renaming for consistency
march_inj_clean <- march_inj %>%
  select(-Problem.Created.Date.Time, -Expected.Return.Date, -Actual.Return.Date, -Days.Out, -Problem.Status, -Reported.Date, -Privacy.Level, -Diagnosis.Code, -Problem.Reported.By) %>%
  rename(ID = Code) %>%
  rename(ProblemDate = Problem.Date) %>%
  rename(BodyPart = Body.Part) %>%
  rename(AffectedArea = Affected.Area) %>%
  rename(Side = Problem.Side) %>%
  rename(InjuryType = Condition) %>%
  rename(BodyPartName = Problem.Sites)

# March: Turning the 'ProblemDate' variable into the proper date format
march_inj_clean$ProblemDate <- mdy_hm(march_inj_clean$ProblemDate)
march_inj_clean$ProblemDate <- as.Date(march_inj_clean$ProblemDate)
str(march_inj_clean$ProblemDate)

# April: Removing unnecessary columns and renaming for consistency
april_inj_clean <- april_inj %>%
  select(-Problem.Created.Date.Time, -Expected.Return.Date, -Actual.Return.Date, -Days.Out, -Problem.Status, -Reported.Date, -Privacy.Level, -Diagnosis.Code, -Problem.Reported.By, -Problem.Reported.By.Provider.Identifier, -Datalys.Injury.Type) %>%
  rename(ID = Code) %>%
  rename(ProblemDate = Problem.Date) %>%
  rename(BodyPart = Body.Part) %>%
  rename(AffectedArea = Affected.Area) %>%
  rename(Side = Problem.Side) %>%
  rename(InjuryType = Condition) %>%
  rename(BodyPartName = Problem.Sites)

# April: Turning the 'ProblemDate' variable into the proper date format
april_inj_clean$ProblemDate <- mdy_hm(april_inj_clean$ProblemDate)
april_inj_clean$ProblemDate <- as.Date(april_inj_clean$ProblemDate)
str(april_inj_clean$ProblemDate)

# May: Removing unnecessary columns and renaming for consistency
may_inj_clean <- may_inj %>%
  select(-Problem.Created.Date.Time, -Expected.Return.Date, -Actual.Return.Date, -Days.Out, -Problem.Status, -Reported.Date, -Privacy.Level, -Diagnosis.Code, -Problem.Reported.By, -Problem.Reported.By.Provider.Identifier, -Datalys.Injury.Type, -Problem.Description) %>%
  rename(ID = Teams) %>%
  rename(ProblemDate = Problem.Date) %>%
  rename(BodyPart = Body.Part) %>%
  rename(AffectedArea = Affected.Area) %>%
  rename(Side = Problem.Side) %>%
  rename(InjuryType = Condition) %>%
  rename(BodyPartName = Problem.Sites)

# May: Turning the 'ProblemDate' variable into the proper date format
may_inj_clean$ProblemDate <- mdy_hm(may_inj_clean$ProblemDate)
may_inj_clean$ProblemDate <- as.Date(may_inj_clean$ProblemDate)
str(may_inj_clean$ProblemDate)
```

# 2. Combining the cleaned datasets prior to filtering for specific injury types
```{r}
combined_inj <- bind_rows(aug_inj_clean, sep_inj_clean, oct_inj_clean, nov_inj_clean, jan_inj_clean, march_inj_clean, april_inj_clean, may_inj_clean)

# checking format & combination accuracy
head(combined_inj, 10)
tail(combined_inj, 10)
dim(combined_inj)
```

# 3. Filtering the data to exclude injuries that may not be predicted w/ guidance from Coach Scott as to what to remove
```{r}
# seeing the different 'InjuryType' options to determine which to keep
unique(combined_inj$InjuryType)

# defining the injury types to exclude
exclude <- c("Contusion", "Insect Bite", "Other Nervous System Disorder", "Other Gastrointestinal Disorder",
             "Asthma", "Other Cardiovascular Disorder", "Mononucleosis", 
             "Pneumonia", "Kidney Injury", "Influenza", "Allergic Reaction", "Other", "Other Injury to Head", "Scheuermann's Disease", "Complaining of pain in toe (finding)", "Cramps")
```


```{r}
# removing these injuries from the data
filtered_inj <- combined_inj %>%
  filter(!InjuryType %in% exclude)

# seeing the different 'InjuryType' options after filtering
unique(filtered_inj$InjuryType)
dim(filtered_inj)
```

# 4. Writing the final injury data to a .csv to be combined with the force plate data 
```{r}
write.csv(filtered_inj, "injury_data.csv", row.names = FALSE)
```

